Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-29 vertical-thermal-review.Rmd
add numbering system to section headers
Update vertical-thermal-review.Rmd

# Introduction

@@

Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests. 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$;
(4) leaf gas exchange, including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) tree and ecosystem ecology.
We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.

@@

# Review of vertical gradients

## The biophysical environment

@@

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "Leaf temperature").

Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].
This holds true across the range of forest types [Fig. `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020]. 
Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "Leaf temperature").

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section "Leaf temperature").

@@

## Leaf temperature

@@

## Leaf traits

@@

@@

## Leaf gas exchange and its thermal sensitivity

@@

### VOC production

@@

## Tree ecology

@@

# Implications

@@

## Conclusions

>>>>

# I. Introduction

@@

Here, we review how the biophysical environment and plant form and function vary across the vertical canopy gradient in forests. 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) leaf temperature ($T_{leaf}$);
(3) the leaf traits that most strongly influence $T_{leaf}$;
(4) leaf gas exchange (including stomatal and boundary layer conductance, photosynthesis, respiration, and volatile organic compound emission) and its thermal sensitivity; and 
(5) tree and ecosystem ecology.
We then consider the implications for understanding forest responses to global change, including how these responses scale across space and time.

@@

# II. Review of vertical gradients

## 1. The biophysical environment

@@

Light, specifically the proportion of incident light and photosynthetically active radiation (PAR), decreases from the canopy top to the forest floor, with the profile shape modified by leaf area density, canopy height, and canopy structure across species and forest types [Fig. `r fig_NEON_vertical`a-d, Supporting Information Figure `r SIfig_NEON_all`, @koike_leaf_2001].
This gradient is more pronounced in dense canopies, including broad-leaf and mixed forests (e.g., SCBI, SERC and HARV, Fig. `r fig_NEON_vertical`d), than in open canopy forests, including many conifer forests [e.g., OSBS, PUUM, and WREF, Fig. `r fig_NEON_vertical`d, Supporting Information Figure `r SIfig_NEON_all`, @chazdonPhotosyntheticLightEnvironments1984; @aussenacInteractionsForestStands2000a; @smithSeasonalDroughtrelatedChanges2019; @tymenQuantifyingMicroenvironmentalVariation2017;  @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. 
Variability in the light environment decreases with height, with the shaded understory being highly influenced by sunflecks, or brief increases in solar radiation caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. 
The implication of this light gradient is that upper canopy leaves have higher potential photosynthetic rate (see section "4. Leaf gas exchange and its thermal sensitivity"), but also greater propensity towards elevated $T_{leaf}$ (see section "2. Leaf temperature").

Wind speeds are also higher at the top of the canopy, owing to the buffering effect of the canopy [@juckerCanopyStructureTopography2018].
This holds true across the range of forest types [Fig. `r fig_NEON_vertical`e, Supporting Information Figure `r SIfig_NEON_all`, @barnardSeasonalVariationCanopy2016a; @mcgregorTreeHeightLeaf2021; @juckerCanopyStructureTopography2018; @mullerEvidenceEfficientNonevaporative2021; @hanberryOpenForestEcosystems2020a; @hanberryReconceptualizationOpenOak2018] and savannas [@curtis_intracanopy_2019;@johnstonWhatLiesVertical2020]. 
Specifically, averaged across the NEON sites, maximum daily wind speeds were 4.1 m s^-1^ at the top of the vertical profile, compared to  $\le$ 1.1 m s^-1^ at the bottom of the vertical profile (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
The implication is that upper canopy leaves have substantially higher boundary layer conductance ($g_b$) and therefore greater potential for both water and heat loss (see section "2. Leaf temperature").

@@

Relative humidity (RH) tends to be higher in the understory and decreases with height, although this trend is absent in open forests [Fig. `r fig_NEON_vertical`f, Supporting Information Figure `r SIfig_NEON_all`, e.g., @juckerCanopyStructureTopography2018; @mcgregorTreeHeightLeaf2021].
Dense-canopy forests maintain higher daily maximum RH in the understory than open forests and nearby open areas, an effect that is greater in wetter conditions, and warmer months [@vonarxSpatiotemporalEffectsForest2012a; @hanberryOpenForestEcosystems2020a]. 
In combination with $T_{air}$, RH determines vapor pressure deficit (VPD)-- the driving force of water loss from leaves--which tends to be greater in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Especially when canopy height and LAI are large, maximum $T_{air}$ and VPD are reduced in the understory [@juckerCanopyStructureTopography2018].
The implication is that canopy leaves face higher $g_b$ and associated potential for water and heat loss (see section "2. Leaf temperature").

@@

## 2. Leaf temperature

@@

## 3. Leaf traits

@@

*beta carotene and lutein are part of the caratenoids pool that responsible for light absorption, and also heat energy dissipation*

@@

## 3. Leaf gas exchange and its thermal sensitivity

@@

### VOC emissions

@@

## 4. Tree and ecosystem ecology

@@

# III.  Implications

@@

## IV. Conclusions
